---
title: "Site selector data visualization tool"
author: "Molly 'Caz' Bell and A.J. Brown using R version 4.2.2"
date: "`r Sys.Date()`"
output: 
  html_document:
    output_dir: "site_selector_temporary_results"
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---
The following is an interactive data report containing water quality data from 
samples taken by the Colorado State University (CSU) Agriculture Water Quality 
Program (AWQP) at a location selected by the user when running the script
(reported below). For more information, please contact A.J. Brown at
Ansley.Brown@colostate.edu or visit the CSU AWQP Website at 
https://waterquality.colostate.edu.

**Please Note:** All analytes are reported in mg/L *EXCEPT*
selenium (reported in ug/L), specific electrical conductivity (EC; mS/cm), and
pH (log H).

## Special Notes

*	All “0” values reported are non-detects (ND) reported by ALS
*	Total suspended solids (TSS), Electrical Conductivity (EC), and pH are performed by AWQP staff at CSU, whereas all other analytes are reported by ALS Laboratories
*	Methods used to derive each analyte can be found in the Raw Data table in the “method” column

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# set working dir
knitr::opts_knit$set(root.dir = here::here())
   # sets the working directory to where the project file is
```

```{r, include=FALSE}
# load required packages
package.list <- c('plotly',
                  'dplyr',
                  'tidyr',
                  'DT',
                  'ggplot2'
                  # 'gridExtra',
                  # 'GGally',
                  # 'PerformanceAnalytics'
                  )
packageLoad <- function(packages){
  for (i in packages) {
    if (!require(i, character.only = TRUE)) {
      install.packages(i, repos = "http://cran.us.r-project.org")
      library(i, character.only = TRUE)
    }
  }
}
packageLoad(package.list)
```

```{r, include=FALSE}
# Uploading file-merge code to create data frame
source('./Code/file-merger.R')
# create data frame
  # note that directory is defined in file_merger.R file
  # change data file path in that script to modify the data source
dat <- returnAllFiles(d = directory, export = TRUE)
dat2 <- dat
# make columns lowercase so they are all the same case
colnames(dat) <- tolower(names(dat))
unique(dat$location.name)
```

```{r, include=FALSE}
# Site list
  # To select site, un-comment your location of interest (alphabetical order)
  # This list may need to be updated as more sites are added/removed
 loc = 'ARDEC' #temporarily ARDEC 2200 until I fix file-merger.r
# loc = 'AVRC STAR'
# loc = 'Barley'
# loc = 'Berthoud'
# loc = 'Big Hollow'
# loc = 'Boulder Lake'
# loc = 'Below Stagecoach Dam'
# loc = 'Gunnison'
# loc = 'Kerbel'
# loc = 'Legacy'
# loc = 'Molina'
# loc = 'Morrison Creek'
# loc = 'Stage Coach Above'
# loc = 'Stage Coach In'
# loc = 'Stagecoach'
# loc = 'The Ranch'
# loc = 'Upper Yampa'
# loc = 'Yellow Jacket'

#create a filtered df to use for violin plots and scatterplot
filter_dataframe <- function(df, selected_location, include_blanks = FALSE) {
  if (include_blanks) {
    filtered_df <- df %>% filter(location.name == selected_location)
  } else {
    filtered_df <- df %>% filter(location.name == selected_location & event.type != "Lab Blank")
  }
  return(filtered_df)
}

# Create df and exclude blanks by default:
filtered_df <- filter_dataframe(dat, loc, include_blanks = FALSE)
```


## Location Selected: <span style="color:#4CAF50">`r loc`</span>

```{r, include=FALSE}
# function to save site data as csv
csv_data <- function(df, location_name) {
  location_data <- df %>%
    filter(location.name == location_name)
  #create output folder
  folder_name <- "site_selector_temporary_results"
  if (!file.exists(folder_name)) {
    dir.create(folder_name)
  }
  
  file_name <- paste0(folder_name, "/", location_name, "_data_", format(Sys.Date(), "%Y-%m-%d"), ".csv")
  write.csv(location_data, file = file_name, row.names = FALSE)
  return(file_name)
}
# use the function to save
file = csv_data(filtered_df, loc)
```

```{r}
# Creating some subset data for static graphing later
analytes_plot1 <- c("Nitrogen, Nitrate (As N)", 
                    "Nitrogen, Nitrite  (As N)", # note the extra space before "(" due to ALS formatting
                    "Nitrogen, Total Kjeldahl")
analytes_plot2 <- c("Phosphorus, Total (As P)", 
                    "Phosphorus, Total Orthophosphate (as P)",
                    "Suspended Solids (Residue, Non-Filterable)")
analytes_plot3 <- setdiff(unique(filtered_df$analyte), c(analytes_plot1, analytes_plot2))
N_df <- filtered_df %>% filter(analyte %in% analytes_plot1)
P_df <- filtered_df %>% filter(analyte %in% analytes_plot2)
Misc_df <- filtered_df %>% filter(analyte %in% analytes_plot3)
```

### CSV Export of selected site data saved as: 
`r file`

# Water Analytes Present and Methods Used for Determination:
```{r}
unique_analyte_method <- filtered_df %>%
  select(analyte, method) %>%
  distinct(analyte, method) %>%
  arrange(analyte)

# Print the table
datatable(unique_analyte_method, 
              caption = "Unique Analytes and Corresponding Methods")

```


# Interactive Violin Plot
Use the below boxplot to graph various water analytes stratified by sample type
(i.e., Inflow, Outflow, etc.). Analytes can be selected by using the dropdown 
menu on the top left of the graph. Hover over points with your mouse for
additional information.

**PLEASE NOTE**: This initial graph will be incorrect (ongoing bug), use the
dropdown menu to select a different analyte first, then go back to the initial
analyte to see it displayed correctly.

```{r, warning=FALSE}
# Create the violin plot function
create_violinplot <- function(df, selected_location) {
  
  tryCatch({
    # Call the function to filter the Data
    
    y_axis_var_names <- sort(unique(df$analyte))
    
    # Specify event colors and types
    event_colors <- c("Inflow" = "#E69F00", "Outflow" = "#56B4E9", "Point Sample" = "#009E73")
    event_types <- c("Inflow", "Outflow", "Point Sample")
    
    # Create the buttons for the dropdown menu
    dropdown_buttons <- lapply(unique(y_axis_var_names), function(analyte_name) {
      y_data <- df$result[df$analyte == analyte_name]
      y_range <- range(c(0, y_data), na.rm = TRUE)
      
      list(
        method = 'restyle',
        args = list('y', list(y_data)),
        label = analyte_name,
        args2 = list('yaxis.range', y_range)  # Update y-axis range
      )
    })
    
    location_violin_plot <- plot_ly(data = df,
                                    x = ~event.type,
                                    y = ~result,
                                    color = ~event.type,
                                    colors = event_colors,
                                    type = 'violin', box = list(
                                      visible = TRUE
                                    ),
                                    meanline = list(
                                      visible = TRUE
                                    )) %>%
      layout(
        title = paste("Violin plots for Location:", selected_location),
        xaxis = list(title = "Event Type"),  # Update x-axis title
        yaxis = list(title = "Result"),
        showlegend = TRUE,
        updatemenus = list(
          list(
            buttons = dropdown_buttons
          )
        )
      )
    
    return(location_violin_plot)
    
  }, error = function(e) {
    cat("An error occurred:", conditionMessage(e))
    return(NULL)
  })
}

# Call the create_violinplot function with your data and selected location
create_violinplot(filtered_df, loc)

```
**To save your plot as a PNG:** hover over the top right corner and click
the camera icon.  The PNG will be saved in your downloads folder.


## Interactive Scatterplot Time Series
Use the below boxplot to graph various water analytes stratified by sample type
(i.e., Inflow, Outflow, etc.) as a time series. Analytes can be selected by 
using the dropdown menu on the top left of the graph. Hover over points with 
your mouse for additional information.

**PLEASE NOTE**: This initial graph will be incorrect (ongoing bug), use the
dropdown menu to select a different analyte first, then go back to the initial
analyte to see it displayed correctly.

```{r}
# Function to create a scatter plot with dropdown
create_scatterplot <- function(df, selected_location) {
  
  tryCatch({
    y_axis_var_names <- sort(unique(df$analyte))
  
    # Specify event colors and types
    event_colors <- c("Inflow" = "#E69F00", "Outflow" = "#56B4E9", "Point Sample" = "#009E73")
    event_types <- c("Inflow", "Outflow", "Point Sample")
    
    create_buttons <- function(y_axis_var_name) {
      analyte_data <- df[df$analyte == y_axis_var_name, ]
      y_data <- analyte_data$result
      
      y_range <- range(y_data)
      buffer_factor <- 0.1  # Set your desired buffer factor
      y_range_with_buffer <- y_range + diff(y_range) * c(-buffer_factor, buffer_factor)
      
      list(
        method = 'restyle',
        args = list('y', list(y_data)),
        label = y_axis_var_name,
        args2 = list('yaxis.range', y_range_with_buffer)  # Update y-axis range
      )
    }
    
    # Create a scatter plot with the first analyte's data
    first_analyte <- y_axis_var_names[1]
    first_analyte_data <- df[df$analyte == first_analyte, ]
    first_analyte_y_range <- create_buttons(first_analyte)
    
    initial_y_data <- first_analyte_data$result
    initial_y_range <- range(initial_y_data)
    buffer_factor <- 0.1
    initial_y_range_with_buffer <- initial_y_range + diff(initial_y_range) * c(-buffer_factor, buffer_factor)
    
    location_scatter_plot <- plot_ly(data = df, x = ~collected, y = ~result,
                                     color = ~event.type, colors = event_colors,
                                     type = 'scatter', mode = 'markers',
                                     marker = list(size = 8, opacity = 0.6),
                                     legendgroup = ~event.type,
                                     showlegend = TRUE) %>%
      layout(
        title = paste("Scatter plot for Location:", selected_location),
        xaxis = list(title = "Collected", tickformat = "%m/%d/%y"),  # Format the date as MM/DD/YY
        yaxis = list(title = "Result", rangeslider = list()),  # Add rangeslider
        showlegend = TRUE,
        updatemenus = list(
          list(
            buttons = lapply(y_axis_var_names, create_buttons)
          )
        )
      )
    
    return(location_scatter_plot)
    
  }, error = function(e) {
    cat("An error occurred:", conditionMessage(e))
    return(NULL)
  })
}

create_scatterplot(filtered_df, loc)

```
**To save your plot as a PNG:** hover over the top right corner and click
the camera icon.  The PNG will be saved in your downloads folder.


# Scatterplot Matrix of All Numerical Analyte Data
Use the below scatter matrix to look at potential relationships between
numerical variables.

```{r, warning=FALSE}
create_scatterplot_matrix <- function(df, location_name) {
  
  tryCatch({
    #special filtering for scatterplot matrix, different from above
    filtered_data <- df %>%
      mutate(sample.id = gsub("-[1-5]$", "", sample.id), # remove -1 to -5 at the end of strings 
             sample.id = gsub("-[1-5]-D$", "-D", sample.id),  # do the same for duplicates
             sample.id = gsub("-2,3,5", "", sample.id)) %>%  # remove special case of "-2,3,5"
      select(sample.id, result, analyte, event.type)

    # Pivot the data wider with aggregation of duplicates by mean
    pivoted_data <- filtered_data %>%
      tidyr::pivot_wider(
        names_from = analyte,
        values_from = result,
        values_fn = list(result = mean)  # Specify how to aggregate duplicates
      )

    # raise error if no data provided
    if (nrow(pivoted_data) == 0) {
      stop("No data available for the given location.")
    }

    # Exclude the first two columns (sample.id and event.type)
    analyte_columns <- colnames(pivoted_data)[-c(1, 2)]  

    # Create scatterplot matrix using pairs()
    pairs(pivoted_data[, analyte_columns],
          labels = analyte_columns,
          col = as.numeric(factor(pivoted_data$event.type)),
          pch = 19)

    # TODO: create scatter matrix using PerformanceAnalytics package
    # chart.Correlation(pd2,
    #                 histogram = T,
    #                 method = 'pearson',
    #                 pch = 8
    #                 )
  }, error = function(e) {
    cat("An error occurred when producing the scatter matrix:",
        conditionMessage(e))
  })
}

create_scatterplot_matrix(filtered_df, loc)
```

# Static Box and Whisker Plots
```{r, fig.height=20, fig.width=12}
create_and_save_boxwhisker_plot <- function(df, output_dir, file_name, plot.title) {
  tryCatch({
    # Ensure the output directory exists
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    
    # Create the file path
    file_path <- file.path(output_dir, file_name)
    
    # Generate the box and whisker plot
    bw_plot <- ggplot(df, aes(x = factor(event.type), y = result, fill = factor(event.type))) +
      geom_boxplot(#outlier.shape = NA, # Hide outliers for clarity
                   position = position_dodge(0)
                   ) + 
      geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) + # Add jitter to show individual data points
      facet_grid(analyte ~ event.count, scales = "free_y", space = "free_x") +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14), # Increase x axis text size
        axis.text.y = element_text(size = 14), # Increase y axis text size
        axis.title = element_text(size = 16), # Increase axis title text size
        strip.text = element_text(size = 16), # Increase facet strip text size
        legend.text = element_text(size = 14), # Increase legend text size
        legend.title = element_text(size = 16), # Increase legend title size
        plot.title = element_text(size = 20, hjust = 0.5), # Increase plot title size
        plot.subtitle = element_text(size = 18), # Increase plot subtitle size
        plot.caption = element_text(size = 12), # Increase plot caption size
        strip.background = element_rect(colour = "black", fill = "white"),
        panel.grid.major = element_line(size = 1), # Increase major grid line size
        panel.grid.minor = element_line(size = 0.5) # Increase minor grid line size
        )+
      labs(title = plot.title, x = "Event Type", y = "Result", fill = "Event Type") +
      scale_fill_brewer(palette = "Pastel1") # Colorblind-friendly palette
    
    # Save the plot as a PNG file
    ggsave(file_path, bw_plot, width = 8, height = 12, units = "in")
    
    # Return the plot object for rendering in the HTML document
    return(bw_plot)
  }, error = function(e) {
    message("An error occurred in create_and_save_boxwhisker_plot: ", e$message)
    return(e$message) # Return error message
  })
}

bw_N <- create_and_save_boxwhisker_plot(N_df,
                                        output_dir = "./site_selector_temporary_results",
                                        file_name = "bw_N.jpg",
                                        plot.title = 'Forms of Nitrogen')
bw_P <- create_and_save_boxwhisker_plot(P_df,
                                        output_dir = "./site_selector_temporary_results",
                                        file_name = "bw_P.jpg",
                                        plot.title = 'Forms of Phosphorous')
bw_Misc <- create_and_save_boxwhisker_plot(Misc_df,
                                            output_dir = "./site_selector_temporary_results",
                                            file_name = "bw_Misc.jpg",
                                        plot.title = 'Water Quality Analytes')

bw_N
bw_P
bw_Misc

```

# Static Bar Plots
```{r, fig.height=20, fig.width=12}
create_and_save_bar_plot <- function(df, output_dir, file_name, plot.title) {
  tryCatch({
    # Ensure the output directory exists
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    
    # Summarize the data
    summary_df <- df %>%
      group_by(event.type, analyte, event.count) %>%
      summarize(mean = mean(result, na.rm = TRUE),
                sd = sd(result, na.rm = TRUE),
                .groups = 'drop')

    # Create the file path
    file_path <- file.path(output_dir, file_name)
    
    # Generate the bar plot with error bars
    bar_plot <- ggplot(summary_df, aes(x = factor(event.type), y = mean, fill = factor(event.type))) +
      geom_bar(stat = "identity", position = position_dodge(0), color = 'black') +
      geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, position = position_dodge(0)) +
      facet_grid(analyte ~ event.count, scales = "free_y", space = "free_x") +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14), # Increase x axis text size
        axis.text.y = element_text(size = 14), # Increase y axis text size
        axis.title = element_text(size = 16), # Increase axis title text size
        strip.text = element_text(size = 16), # Increase facet strip text size
        legend.text = element_text(size = 14), # Increase legend text size
        legend.title = element_text(size = 16), # Increase legend title size
        plot.title = element_text(size = 20, hjust = 0.5), # Increase plot title size
        plot.subtitle = element_text(size = 18), # Increase plot subtitle size
        plot.caption = element_text(size = 12), # Increase plot caption size
        strip.background = element_rect(colour = "black", fill = "white"),
        panel.grid.major = element_line(size = 1), # Increase major grid line size
        panel.grid.minor = element_line(size = 0.5) # Increase minor grid line size
        )+
      labs(title = plot.title, x = "Event Type", y = "Result", fill = "Event Type") +
      scale_fill_brewer(palette = "Pastel1") # Colorblind-friendly palette

    # Save the plot as a PNG file
    ggsave(file_path, bar_plot, width = 8, height = 12, units = "in")
    
    # Return the plot object for rendering in the HTML document
    return(bar_plot)
  }, error = function(e) {
    message("An error occurred in create_and_save_bar_plot: ", e$message)
    return(e$message) # Return error message
  })
}

bar_N <- create_and_save_bar_plot(N_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_N.jpg",
                                  plot.title = 'Forms of Nitrogen')
bar_P <- create_and_save_bar_plot(P_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_P.jpg",
                                  plot.title = 'Forms of Phosphorus')
bar_Misc <- create_and_save_bar_plot(Misc_df,
                                     output_dir = "./site_selector_temporary_results",
                                     file_name = "bar_Misc.jpg",
                                    plot.title = 'Water Quality Analytes')

bar_N
bar_P
bar_Misc
```


# Raw Data
Use the table below to look at raw values.
```{r}
datatable(filtered_df)
```