---
title: "Site selector data visualization tool"
author: "Molly 'Caz' Bell and A.J. Brown using R version 4.2.2"
date: "`r Sys.Date()`"
output: 
  html_document:
    output_dir: "site_selector_temporary_results"
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---
The following is an interactive data report containing water quality data from 
samples taken by the Colorado State University (CSU) Agriculture Water Quality 
Program (AWQP) at a location selected by the user when running the script
(reported below). For more information, please contact A.J. Brown at
Ansley.Brown@colostate.edu or visit the CSU AWQP Website at 
https://waterquality.colostate.edu.

**Please Note:** All analytes are reported in mg/L *EXCEPT*
selenium (reported in ug/L), specific electrical conductivity (EC; mS/cm), and
pH (log H).

## Special Notes

*	All “0” values reported are non-detects (ND) reported by ALS
*	Total suspended solids (TSS), Electrical Conductivity (EC), and pH are performed by AWQP staff at CSU, whereas all other analytes are reported by ALS Laboratories
*	Methods used to derive each analyte can be found in the Raw Data table in the “method” column

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# set working dir
knitr::opts_knit$set(root.dir = here::here())
   # sets the working directory to where the project file is
```

```{r, include=FALSE}
# load required packages
package.list <- c('plotly',
                  'dplyr',
                  'tidyr',
                  'DT',
                  'ggplot2',
                  'stringr'
                  # 'gridExtra',
                  # 'GGally',
                  # 'PerformanceAnalytics'
                  )
packageLoad <- function(packages){
  for (i in packages) {
    if (!require(i, character.only = TRUE)) {
      install.packages(i, repos = "http://cran.us.r-project.org")
      library(i, character.only = TRUE)
    }
  }
}
packageLoad(package.list)
```

```{r, include=FALSE}
# # Uploading file-merge code to create data frame
source('./Code/file-merger.R')
# # create data frame
#   # note that directory is defined in file_merger.R file
#   # change data file path in that script to modify the data source
# dat <- returnAllFiles(d = directory, export = TRUE)

# import 2200 data manually for this report from the python scrip
dat <- read.csv('./Report/merged_data.csv', header = TRUE, stringsAsFactors = FALSE)
dat2 <- dat

```

```{r}
# some more data cleaning prior to filtered df

# create an uppercase copy of sample.id for processData
dat$SAMPLE.ID <- dat$sample.id

# change all 2200 names to 'ARDEC'
dat$location.name[dat$location.name == '2200'] <- 'ARDEC'
unique(dat$location.name)

# run processData again to make sure event.type is correct
dat <- processData(dat)

# drop all event.types that are Point Samples
dat <- dat[dat$event.type != 'Point Sample',]

# make result numeric
dat$result <- as.numeric(dat$result)

# drop all rows where 'result' is NA
dat <- dat[!is.na(dat$result),]

# create new year column based on collected column and make it a factor
dat$year <- format(as.Date(dat$collected, format = "%m/%d/%Y"), "%Y")
```


```{r, include=FALSE}
# Site list
  # To select site, un-comment your location of interest (alphabetical order)
  # This list may need to be updated as more sites are added/removed
 loc = 'ARDEC' #temporarily ARDEC 2200 until I fix file-merger.r
# loc = 'AVRC STAR'
# loc = 'Barley'
# loc = 'Berthoud'
# loc = 'Big Hollow'
# loc = 'Boulder Lake'
# loc = 'Below Stagecoach Dam'
# loc = 'Gunnison'
# loc = 'Kerbel'
# loc = 'Legacy'
# loc = 'Molina'
# loc = 'Morrison Creek'
# loc = 'Stage Coach Above'
# loc = 'Stage Coach In'
# loc = 'Stagecoach'
# loc = 'The Ranch'
# loc = 'Upper Yampa'
# loc = 'Yellow Jacket'

#create a filtered df to use for violin plots and scatterplot
filter_dataframe <- function(df, selected_location, include_blanks = FALSE) {
  if (include_blanks) {
    filtered_df <- df %>% filter(location.name == selected_location)
  } else {
    filtered_df <- df %>% filter(location.name == selected_location & event.type != "Lab Blank")
  }
  return(filtered_df)
}

# Create df and exclude blanks by default:
filtered_df <- filter_dataframe(dat, loc, include_blanks = FALSE)
```


## Location Selected: <span style="color:#4CAF50">`r loc` 2200</span>

```{r, include=FALSE}
# function to save site data as csv
csv_data <- function(df, location_name) {
  location_data <- df %>%
    filter(location.name == location_name)
  #create output folder
  folder_name <- "site_selector_temporary_results"
  if (!file.exists(folder_name)) {
    dir.create(folder_name)
  }
  
  file_name <- paste0(folder_name, "/", location_name, "_data_", format(Sys.Date(), "%Y-%m-%d"), ".csv")
  write.csv(location_data, file = file_name, row.names = FALSE)
  return(file_name)
}
# use the function to save
file = csv_data(filtered_df, loc)
```

```{r}
# Abbreviating analyte names
filtered_df <- filtered_df %>%
  mutate(analyte_abbrev = case_when(
    analyte == "Nitrogen, Nitrate (As N)" ~ "NO3-N",
    analyte == "Nitrogen, Total Kjeldahl" ~ "TKN",
    analyte == "Phosphorus, Total (As P)" ~ "P-Total",
    analyte == "Selenium, Total" ~ "Se",
    analyte == "Phosphorus, Total Orthophosphate (as P)" ~ "Ortho-P",
    analyte == "Nitrogen, Nitrite  (As N)" ~ "N-NO2",
    analyte == "Total Dissolved Solids (Residue, Filterable)" ~ "TDS",
    analyte == "pH" ~ "pH",
    analyte == "Suspended Solids (Residue, Non-Filterable)" ~ "TSS",
    analyte == "Specific Conductance" ~ "EC",
    TRUE ~ analyte # Default case to keep original name if no match
  ))

```


```{r}
# Creating some subset data for static graphing later
analytes_plot1 <- c("Nitrogen, Nitrate (As N)", 
                    "Nitrogen, Nitrite  (As N)", # note the extra space before "(" due to ALS formatting
                    "Nitrogen, Total Kjeldahl")
analytes_plot2 <- c("Phosphorus, Total (As P)", 
                    "Phosphorus, Total Orthophosphate (as P)",
                    "Suspended Solids (Residue, Non-Filterable)")
analytes_plot3 <- setdiff(unique(filtered_df$analyte), c(analytes_plot1, analytes_plot2))
N_df <- filtered_df %>% filter(analyte %in% analytes_plot1)
P_df <- filtered_df %>% filter(analyte %in% analytes_plot2)
Misc_df <- filtered_df %>% filter(analyte %in% analytes_plot3)
```

### CSV Export of selected site data saved as: 
`r file`

# Water Analytes Present and Methods Used for Determination:
```{r}
unique_analyte_method <- filtered_df %>%
  select(analyte, method) %>%
  distinct(analyte, method) %>%
  arrange(analyte)

# Print the table
datatable(unique_analyte_method, 
              caption = "Unique Analytes and Corresponding Methods")

```


# Static Box and Whisker Plots
```{r, fig.height=20, fig.width=12}
create_and_save_boxwhisker_plot <- function(df, output_dir, file_name, plot.title) {
  tryCatch({
    # Ensure the output directory exists
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    
    # Create the file path
    file_path <- file.path(output_dir, file_name)
    
    # Generate the box and whisker plot
    bw_plot <- ggplot(df, aes(x = factor(event.type), y = result, fill = factor(event.type))) +
      geom_boxplot(#outlier.shape = NA, # Hide outliers for clarity
                   position = position_dodge(0)
                   ) + 
      geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) + # Add jitter to show individual data points
      facet_grid(analyte_abbrev ~ event.count, scales = "free_y", space = "free_x") +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14), # Increase x axis text size
        axis.text.y = element_text(size = 14), # Increase y axis text size
        axis.title = element_text(size = 16), # Increase axis title text size
        strip.text = element_text(size = 16), # Increase facet strip text size
        legend.text = element_text(size = 14), # Increase legend text size
        legend.title = element_text(size = 16), # Increase legend title size
        plot.title = element_text(size = 20, hjust = 0.5), # Increase plot title size
        plot.subtitle = element_text(size = 18), # Increase plot subtitle size
        plot.caption = element_text(size = 12), # Increase plot caption size
        strip.background = element_rect(colour = "black", fill = "white"),
        panel.grid.major = element_line(size = 1), # Increase major grid line size
        panel.grid.minor = element_line(size = 0.5) # Increase minor grid line size
        )+
      labs(title = plot.title, x = "Event Type", y = "Result", fill = "Event Type") +
      scale_fill_brewer(palette = "Pastel1") # Colorblind-friendly palette
    
    # Save the plot as a PNG file
    ggsave(file_path, bw_plot, width = 12, height = 8, units = "in")
    
    # Return the plot object for rendering in the HTML document
    return(bw_plot)
  }, error = function(e) {
    message("An error occurred in create_and_save_boxwhisker_plot: ", e$message)
    return(e$message) # Return error message
  })
}

bw_N <- create_and_save_boxwhisker_plot(N_df,
                                        output_dir = "./site_selector_temporary_results",
                                        file_name = "bw_N.jpg",
                                        plot.title = 'Forms of Nitrogen')
bw_P <- create_and_save_boxwhisker_plot(P_df,
                                        output_dir = "./site_selector_temporary_results",
                                        file_name = "bw_P.jpg",
                                        plot.title = 'Forms of Phosphorous')
bw_Misc <- create_and_save_boxwhisker_plot(Misc_df,
                                            output_dir = "./site_selector_temporary_results",
                                            file_name = "bw_Misc.jpg",
                                        plot.title = 'Water Quality Analytes')

bw_N
bw_P
bw_Misc

```

# Static Bar Plots
```{r, fig.height=20, fig.width=12}
create_and_save_bar_plot <- function(df, output_dir, file_name, plot.title) {
  tryCatch({
    # Ensure the output directory exists
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    
    # Summarize the data
    summary_df <- df %>%
      group_by(event.type, analyte_abbrev, event.count) %>%
      summarize(mean = mean(result, na.rm = TRUE),
                sd = sd(result, na.rm = TRUE),
                .groups = 'drop')

    # Create the file path
    file_path <- file.path(output_dir, file_name)
    
    # Generate the bar plot with error bars
    bar_plot <- ggplot(summary_df, aes(x = factor(event.type), y = mean, fill = factor(event.type))) +
      geom_bar(stat = "identity", position = position_dodge(0), color = 'black') +
      geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, position = position_dodge(0)) +
      facet_grid(analyte_abbrev ~ event.count, scales = "free_y", space = "free_x") +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14), # Increase x axis text size
        axis.text.y = element_text(size = 14), # Increase y axis text size
        axis.title = element_text(size = 16), # Increase axis title text size
        strip.text = element_text(size = 16), # Increase facet strip text size
        legend.text = element_text(size = 14), # Increase legend text size
        legend.title = element_text(size = 16), # Increase legend title size
        plot.title = element_text(size = 20, hjust = 0.5), # Increase plot title size
        plot.subtitle = element_text(size = 18), # Increase plot subtitle size
        plot.caption = element_text(size = 12), # Increase plot caption size
        strip.background = element_rect(colour = "black", fill = "white"),
        panel.grid.major = element_line(size = 1), # Increase major grid line size
        panel.grid.minor = element_line(size = 0.5) # Increase minor grid line size
        )+
      labs(title = plot.title, x = "Event Type", y = "Result", fill = "Event Type") +
      scale_fill_brewer(palette = "Pastel1") # Colorblind-friendly palette

    # Save the plot as a PNG file
    ggsave(file_path, bar_plot, width = 12, height = 8, units = "in")
    
    # Return the plot object for rendering in the HTML document
    return(bar_plot)
  }, error = function(e) {
    message("An error occurred in create_and_save_bar_plot: ", e$message)
    return(e$message) # Return error message
  })
}

bar_N <- create_and_save_bar_plot(N_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_N.jpg",
                                  plot.title = 'Forms of Nitrogen')
bar_P <- create_and_save_bar_plot(P_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_P.jpg",
                                  plot.title = 'Forms of Phosphorus')
bar_Misc <- create_and_save_bar_plot(Misc_df,
                                     output_dir = "./site_selector_temporary_results",
                                     file_name = "bar_Misc.jpg",
                                    plot.title = 'Water Quality Analytes')

bar_N
bar_P
bar_Misc
```

# Static Bar Plots by Year
```{r}
create_and_save_bar_plot_year <- function(df, output_dir, file_name, plot.title) {
  tryCatch({
    # Ensure the output directory exists
    if (!dir.exists(output_dir)) {
      dir.create(output_dir, recursive = TRUE)
    }
    
    # Assuming 'year' column is present in the df
    # If not, you might need to create it from a date column
    
    # Summarize the data
    summary_df <- df %>%
      group_by(year, analyte_abbrev, event.type) %>%
      summarize(mean = mean(result, na.rm = TRUE),
                sd = sd(result, na.rm = TRUE),
                .groups = 'drop')

    # Create the file path
    file_path <- file.path(output_dir, file_name)
    
    # Generate the bar plot with error bars
    bar_plot <- ggplot(summary_df, aes(x = factor(year), y = mean, fill = factor(event.type))) +
      geom_bar(stat = "identity", position = position_dodge(1), color = 'black') + # Adjusted position for side-by-side bars
      geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, position = position_dodge(1)) + # Adjust position for error bars
      facet_grid(analyte_abbrev ~ ., scales = "free_y", space = "free_x") +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14), # Increase x axis text size
        axis.text.y = element_text(size = 14), # Increase y axis text size
        axis.title = element_text(size = 16), # Increase axis title text size
        strip.text = element_text(size = 16), # Increase facet strip text size
        legend.text = element_text(size = 14), # Increase legend text size
        legend.title = element_text(size = 16), # Increase legend title size
        plot.title = element_text(size = 20, hjust = 0.5), # Increase plot title size
        plot.subtitle = element_text(size = 18), # Increase plot subtitle size
        plot.caption = element_text(size = 12), # Increase plot caption size
        strip.background = element_rect(colour = "black", fill = "white"),
        panel.grid.major = element_line(size = 1), # Increase major grid line size
        panel.grid.minor = element_line(size = 0.5) # Increase minor grid line size
        ) +
      labs(title = plot.title, x = "Year", y = "Result", fill = "Event Type") +
      scale_fill_brewer(palette = "Pastel1") # Colorblind-friendly palette

    # Save the plot as a PNG file
    ggsave(file_path, bar_plot, width = 12, height = 8, units = "in")
    
    # Return the plot object for rendering in the HTML document
    return(bar_plot)
  }, error = function(e) {
    message("An error occurred in create_and_save_bar_plot: ", e$message)
    return(e$message) # Return error message
  })
}


bar_N_byYear <- create_and_save_bar_plot_year(N_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_N_year.jpg",
                                  plot.title = 'Forms of Nitrogen')
bar_P_byYear <- create_and_save_bar_plot_year(P_df,
                                  output_dir = "./site_selector_temporary_results",
                                  file_name = "bar_P_year.jpg",
                                  plot.title = 'Forms of Phosphorus')
bar_Misc_byYear <- create_and_save_bar_plot_year(Misc_df,
                                     output_dir = "./site_selector_temporary_results",
                                     file_name = "bar_Misc_year.jpg",
                                    plot.title = 'Water Quality Analytes')

bar_N_byYear
bar_P_byYear
bar_Misc_byYear

```


```{r}
# calculate differences between inflow and outflow values for each analyte, event.count, and year

# calculate summary stats and differences between years
summary_df <- filtered_df %>%
  group_by(year, analyte, event.type) %>%
  summarize(mean_result = mean(result, na.rm = TRUE),
            sd = sd(result, na.rm = TRUE),
            .groups = 'drop')
summary_df

# Creating inflow and outflow DataFrames
inflow_df <- summary_df %>% filter(event.type == "Inflow")
outflow_df <- summary_df %>% filter(event.type == "Outflow")

# Renaming columns in inflow_df for clarity
inflow_df <- inflow_df %>% rename(mean_inflow = mean_result, sd_inflow = sd)

# Renaming columns in outflow_df for clarity
outflow_df <- outflow_df %>% rename(mean_outflow = mean_result, sd_outflow = sd)

# Merging the DataFrames
merged_df <- left_join(inflow_df, outflow_df, by = c("year", "analyte"))

# Calculating the differences
merged_df <- merged_df %>% 
  mutate(mean_difference = mean_inflow - mean_outflow,
         removal_efficiency = (mean_difference / mean_inflow)*100
         )

merged_df
```



# Raw Data
Use the table below to look at raw values.
```{r}
datatable(filtered_df)
```